---
title: "Statistics for figure 10 - granules per chloroplast"
author: Jiawen Chen/Statistics Services Ltd (Reading, UK)
output: rmarkdown::github_document
date: "2022-09-23"
---

We consulted Statistics Services Ltd (Reading, UK) for the analysis of our count data regarding granule per chloroplast, obtained from counting from light microscopy images of Arabidopsis leaf sections. 

We used mixed effects models for this, in order to take into account both the three biological replicates for each genotype (random effect), and the distribution of granule number chloroplast for each genotype. We tried a Poisson model first, but this showed overdispersion so we then proceeded with negative binomial models, which were not overdispersed. 


# Analysis 1: Non-backcrossed lines: WT Kronos, *mrc-1*, *mrc-2*, *mrc-3*

Importing and sorting data
```{r, results=FALSE}
dat<-read.csv("2021_gpc_nonbackcrossed.csv")
dat$rep<-factor(dat$rep)
dat$group<-factor(dat$group)
dat$counts.f<-ordered(dat$counts)
colnames(dat)

#Making a unique rep id column for 3 biological reps in each of 4 genotype groups
dat$rep.id<-dat$rep:dat$group
length(levels(dat$rep.id))
head(dat)
# Disaggregating the data to use for fitting functions
dat.disag<-dat[rep(1:nrow(dat),dat[,2]),-2]
head(dat.disag)
aggregate(dat$freq~dat$rep.id,FUN=sum)
aggregate(dat.disag$counts~dat.disag$rep.id,FUN=length)

```


## Mixed effect models, fitted by maximum likelihood. Negative binomial regression. 
```{r, results=FALSE, message=FALSE, warning=FALSE}
## Default repo
local({r <- getOption("repos")
       r["CRAN"] <- "http://cran.r-project.org" 
       options(repos=r)
})

install.packages("MASS");
library(MASS)
nb0<-glm.nb(counts~1,data=dat.disag) #Must be fitted before loading GLMMadaptive to prevent masking

install.packages("GLMMadaptive")
library(GLMMadaptive); library(emmeans)



```


```{r}
nb1<-mixed_model(counts~1,~1|rep.id,data=dat.disag,family=negative.binomial)
nb2<-mixed_model(counts~group,~1|rep.id,data=dat.disag,family=negative.binomial)
anova(nb1,nb0) #Test of random effect component, resulting p-value must be halved #ignore warning, it's caused by masking
```


We see a significant random effect of the biological replicates. Now we proceed to testing the fixed effect of the genotype adjusted for random effect. 

```{r}
anova(nb1,nb2) 

```

We see a significant fixed effect adjusted for random effect. Now we proceed to means and pairwise comparisons.

```{r}
# Must regrid, or the argument type is ignored
nb2r<-update(ref_grid(nb2),tran='log')
emmeans(nb2r,~group,type='response') # predicted mean count
pairs(emmeans(nb2r,~group,type='response'),inf=T)
pairs(emmeans(nb2r,~group,type='response'),rev=T,inf=T)

# Illustrate fitted conditional distributions
theta<-exp(nb2$phis); theta 

# Note arg size = theta 
curve(dnbinom(x,size=theta,mu=5.21),0,21,22,lwd=2,ylim=c(0,.25),xlab='count',ylab='pmf')
curve(dnbinom(x,size=theta,mu=2.61),0,21,22,lwd=2,add=T,col=2)
curve(dnbinom(x,size=theta,mu=3.75),0,21,22,lwd=2,add=T,col=3)
curve(dnbinom(x,size=theta,mu=4.77),0,21,22,lwd=2,add=T,col=4)
legend('topright',levels(dat$group),lwd=2,col=1:4,title='group:')
title('Fitted conditional Negative Binomial distribution')


```

Finally, we check that the data is not overdispersed. We use an informal goodness of fit criterion, e.g. from [Gbur et al 2012](https://acsess.onlinelibrary.wiley.com/doi/book/10.2134/2012.generalized-linear-mixed-models), section 5.4, page 128.
Sum of conditional Pearson residuals squared/model DF.
A ratio >>1 suggests more observed variance than expected from a Poisson/negative binomial. 
```{r}
dat.disag$pred.ss<-predict(nb2,dat.disag,type='sub')
dat.disag$pearson<-(dat.disag$counts-dat.disag$pred.ss)/sqrt(dat.disag$pred.ss+dat.disag$pred.ss^2/theta)
pearchi2<-sum(dat.disag$pearson^2)
pearchi2/nobs(nb2) 

```
There was overdispersion when we used a Poisson model, but no overdispersion with this negative binomial model.


# Analysis 2: Backcrossed lines: WT Kronos, *mrc-1*, *mrc-1* BC2 *aabb*,  *mrc-1* BC2 AABB
We did the same analysis on these lines, from the file "2021_gpc_backcrossed.csv". Details of analysis identical to above, showing the mean and pairwise comparison outputs below.

```{r, include=FALSE}
dat2<-read.csv("2021_gpc_backcrossed.csv")
colnames(dat2)
dat2$rep<-factor(dat2$rep)
dat2$group<-factor(dat2$group)
dat2$counts.f<-ordered(dat2$counts)
dat2$counts<-as.integer(dat2$counts)
dat2$freq<-as.integer(dat2$freq)

sapply(dat2, class)
dat2$rep.id<-dat2$rep:dat2$group
length(levels(dat2$rep.id))
head(dat2)

# THE FITTING FUNCTION REQUIRES DISAGGREGATED DATA
dat2.disag<-dat2[rep(1:nrow(dat2),dat2[,2]),-2]
head(dat2.disag)
aggregate(dat2$freq~dat2$rep.id,FUN=sum)
aggregate(dat2.disag$counts~dat2.disag$rep.id,FUN=length)
# NEGATIVE BINOMIAL REGRESSION
##############################
library(MASS)
nb02<-glm.nb(counts~1,data=dat2.disag) #MUST BE FITTED BEFORE LOADING GLMMadaptive TO PREVENT MASKING

library(GLMMadaptive); library(emmeans)
nb12<-mixed_model(counts~1,~1|rep.id,data=dat2.disag,family=negative.binomial)
nb22<-mixed_model(counts~group,~1|rep.id,data=dat2.disag,family=negative.binomial)
anova(nb12,nb02) #TEST OF RANDOM EFFECT COMPONENT, RESULTING P-VALUE MUST BE HALVED
#there is an effect

# IGNORE WARNING, IT'S CAUSED BY MASKING
anova(nb12,nb22) #TEST OF FIXED EFFECT ADJUSTED FOR RANDOM EFFECT
#there is an effect

# MUST REGRID, OR THE ARG type IS IGNORED
nb2r2<-update(ref_grid(nb22),tran='log')
```


```{r}
emmeans(nb2r2,~group,type='response') # Predicted mean count
pairs(emmeans(nb2r2,~group,type='response'),inf=T)
pairs(emmeans(nb2r2,~group,type='response'),rev=T,inf=T)
```


```{r, include=FALSE}
# ILLUSTRATE FITTED CONDITIONAL DISTRIBUTIONS
theta2<-exp(nb22$phis); theta2

# INFORMAL GOODNESS OF FIT CRITERION, e.g. FROM GBUR ET AL 2012, SECTION 5.4, PAGE 128
# SUM OF CONDITIONAL PEARSON RESIDUALS SQUARED / MODEL DF
# A RATIO >>1 SUGGESTS MORE OBSERVED VARIANCE THAN EXPECTED FROM A MODEL
dat2.disag$pred.ss<-predict(nb22,dat2.disag,type='sub')
dat2.disag$pearson<-(dat2.disag$counts-dat2.disag$pred.ss)/sqrt(dat2.disag$pred.ss+dat2.disag$pred.ss^2/theta2)
pearchi22<-sum(dat2.disag$pearson^2)
pearchi22/nobs(nb22) 
```
```

